Visualizing probability distributions across bivariate cyclic temporal granularities

### 2021-03-24 reviews/response.Rmd

---
title: Response to reviewers
author: JCGS-20-344
output:
  MonashEBSTemplates::memo:
    includes:
      in_header: preamble2.tex
fontfamily: mathpazo
---

We thank the reviewers for their helpful feedback which has led to many improvements in the paper. The point by point description of changes are below: the reviewers' comments are in red and our response is in black.

This manuscript has undergone an overhaul to address reviewer's comments:

1. 
2. 
3. 
4.

Along with these major changes, we have done lots of minor tweaks to improve the manuscript.

# Reviewer 1 comments {-}

> comment 1.

response

> comment 2

response

> comment 3

response

> comment 4

response

### 2021-03-25 reviews/response.Rmd

---
title: Response to reviewers
author: "JCGS-20-344"
output:
  MonashEBSTemplates::memo:
    includes:
      in_header: preamble2.tex
fontfamily: mathpazo
---

We would really like to thank the editor and the reviewers for their encouragement and detailed feedback which has helped to improve the paper in terms of both concision and precision and as well as reaching out to more readers.
The point by point description of changes are below: the reviewers' comments are in red and our response is in black.

@@

response

> comment 2

response

> I have to confess that I do not know anything about the sport cricket and it may be unthinkable for people from countries where cricket is a national sport, but I really had difficulties to follow the example in Section 6.2. The basic hierarchical concept of over—inning--match--season was understandable, but I think that some parts need to be revisited and clarified for also reaching readers not knowledgeable in cricket. Especially the discussion on the interesting questions raised are a bit difficult to follow without knowing much about cricket. For example some terms should be at least briefly explained or removed if not necessary for the explanation, e.g., Twenty20 format, fielding, bowling, dismissal, wicket. I do know a lot about different sports and also the basic idea and structure in many sports, but cricket was never on my radar. Apologise my ignorance about cricket.

Thanks for the advice. We have added a description of the game cricket in the first paragraph of Section 6.2 so that each new terminology applicable for cricket (and relevant for the following analysis) is highlighted in italics and briefly explained. We believe that this will help the readers not knowledgeable in cricket to also follow the presented analysis.

### 2021-03-25 main.Rmd
cricket description added to address comments

Our proposed approach can be generalized to other hierarchical granularities where there is an underlying ordered index. We illustrate this with data from the sport cricket. Although there is no conventional time component in cricket, each ball can be thought to represent an ordering over the course of the game. In the Twenty20 format, an over will consist of 6 balls (with some exceptions), an innings is restricted to a maximum of 20 overs, a match will consist of 2 innings and a season consists of several matches. Thus, there is a hierarchy where ball is nested within overs, overs nested within innings, and innings within matches. Cyclic granularities can be constructed using this hierarchy. Example granularities include ball of the over, over of the innings, and ball of the innings. The hierarchy table is given in  \autoref{tab:hierarchy-cric}.

@@

Although most of these cyclic granularities are circular by design of the hierarchy, in practice some granularities are aperiodic. For example, most overs will consist of 6 balls, but there are exceptions due to wide balls, no-balls, or when an innings finishes before the over finishes. Thus, the cyclic granularity ball-of-over may be aperiodic.

The Indian Premier League (IPL) is a professional Twenty20 cricket league in India contested by eight teams representing eight different cities in India. The IPL ball-by-ball data is provided in the `cricket` data set in the `gravitas` package for a sample of 214 matches spanning 9 seasons (2008 to 2016) such that each over has 6 balls, each innings has 20 overs and each match has 2 innings.

There are many interesting questions that could be addressed with the `cricket` data set. For example, does the distribution of total runs vary depending on if a team bats in the first or second innings? The Mumbai Indians (MI) and Chennai Super Kings (CSK) appeared in final playoffs from 2010 to 2015. Using  data from these two teams, it can be observed (\autoref{fig:cricex}a) that for the team batting in the first innings there is an upward trend of runs per over, while there is no clear upward trend in median and quartile deviation of runs for the team batting in the second innings after the first few overs. This suggests that players feel mounting pressure to score more runs as they approach the end of the first innings, while teams batting second have a set target in mind and are not subjected to such mounting pressure and therefore may adopt a more conservative run-scoring strategy.

Another question that can be addressed is if good fielding or bowling (defending) in the previous over affects the scoring rate in the subsequent over? To measure the defending quality, we use an indicator function on dismissals (1 if there was at least one wicket in the previous over, 0 otherwise). The scoring rate is measured by runs per over. \autoref{fig:cricex}b shows that no dismissals in the previous over leads to a higher median and quartile spread of runs per over compared to the case when there has been at least one dismissal in the previous over. This seems to be unaffected by the over of the innings (the facet variable). This might be because the new batsman needs to play himself in or the dismissals lead the (not-dismissed) batsman to adopt a more defensive play style. Run rates will also vary depending on which player is facing the next over and when the wicket falls in the previous over.

@@

(ref:cricex) Runs per over shown with different distribution displays, and granularities. Plot a shows letter value plots across overs faceted by innings. For the team batting in the first innings there is an upward trend of runs per over, while there is no such pattern of runs for the teams batting in the second innings. Plot b shows quantile plots of runs per over across an indicator of wickets in the previous over faceted by current over. This indicates that at least one wicket in the previous over leads to lower median run rate and quartile spread in the subsequent over, regardless of the over of the innings.

>>>>

Our proposed approach can be generalized to other hierarchical granularities where there is an underlying ordered index. We illustrate this with data from the sport cricket. Cricket is played with two teams of 11 players each, with each team taking turns batting and playing the field. This is similar to baseball, wherein the _batsman_ and _bowler_ in cricket are analogous to a batter and pitcher in baseball.  A _wicket_ is a structure with three sticks, stuck into the ground, placed at the end of the cricket pitch behind the batsman. A batsman tries to prevent the bowler from hitting the wicket by hitting the _ball_ using a _bat_. Two batsmen of a team are on the pitch at the same time. There are nine _fielders_ on the opposing team, with one player out of the 11 acting as the bowler, while another takes up the role of the _wicket-keeper_. To score a _run_, a batsman can run after hitting the ball and changing places on the pitch or also by hitting the ball out of the field on a fly (6 runs) or on a bounce (4 runs). A batsman is _dismissed_ when the period of batting is brought to an end either due to good bowling of the bowler or good fielding of the rest of the opposing team.

Cricket is played in various formats and Twenty20 cricket (T20) is a shortened format, where the two teams have a single _innings_ each, which is restricted to a maximum of 20 _overs_. An over will consist of 6 balls (with some exceptions). A single _match_ will consist of 2 innings and a _season_ consists of several matches. Although there is no conventional time component in cricket, each ball can be thought to represent an ordering over the course of the game. Then, we can conceive a hierarchy where the ball is nested within overs, overs nested within innings, innings within matches, and matches within seasons. Cyclic granularities can be constructed using this hierarchy. Example granularities include ball of the over, over of the innings, and ball of the innings. The hierarchy table is given in  \autoref{tab:hierarchy-cric}. Although most of these cyclic granularities are circular by the design of the hierarchy, in practice some granularities are aperiodic. For example, most overs will consist of 6 balls, but there are exceptions due to wide balls, no-balls, or when an innings finishes before the over finishes. Thus, the cyclic granularity ball-of-over may be aperiodic.

@@

The Indian Premier League (IPL) is a professional T20 cricket league in India contested by eight teams representing eight different cities in India. The IPL ball-by-ball data is provided in the `cricket` data set in the `gravitas` package for a sample of 214 matches spanning 9 seasons (2008 to 2016) such that each over has 6 balls, each innings has 20 overs and each match has 2 innings.

Many interesting questions could be addressed with the `cricket` data set. For example, does the distribution of total runs vary depending on if a team bats in the first or second innings? The Mumbai Indians (MI) and Chennai Super Kings (CSK) appeared in the final playoffs from 2010 to 2015. Using data from these two teams, it can be observed (\autoref{fig:cricex}a) that for the team batting in the first innings there is an upward trend of runs per over, while there is no clear upward trend in the median and quartile deviation of runs for the team batting in the second innings after the first few overs. This suggests that players feel mounting pressure to score more runs as they approach the end of the first innings, while teams batting second have a set target in mind and are not subjected to such mounting pressure and therefore may adopt a more conservative run-scoring strategy.

Another question that can be addressed is if good fielding or bowling (defending) in the previous over affects the scoring rate in the subsequent over? To measure the defending quality, we use an indicator function on dismissals (1 if there was at least one wicket in the previous over, 0 otherwise). The scoring rate is measured by runs per over. \autoref{fig:cricex}b shows that no dismissals in the previous over leads to a higher median and quartile spread of runs per over compared to the case when there has been at least one dismissal in the previous over. This seems to be unaffected by the over of the innings (the faceting variable). This might be because the new batsman needs to play himself in or the dismissals lead the (not-dismissed) batsman to adopt a more defensive playstyle. Run rates will also vary depending on which player is facing the next over and when the wicket falls in the previous over.

@@

(ref:cricex) Runs per over shown with different distribution displays, and granularities. Plot a shows letter value plots across overs faceted by innings. For the team batting in the first innings, there is an upward trend of runs per over, while there is no such pattern of runs for the teams batting in the second innings. Plot b shows quantile plots of runs per over across an indicator of wickets in the previous over faceted by current over. This indicates that at least one wicket in the previous over leads to a lower median run rate and quartile spread in the subsequent over, regardless of the over of the innings.
